---
title: "TDC protein velocity"
output: html_document
date: "2025-08-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load necessary libraries 
```{r}
library(tidyverse)
library(survival)
library(broom)
```


Prep the data 
```{r}
# load cancer, protein, and followup time data
cancer_data <- read_csv("cancer_data.csv")
proteomic_data <- read_csv("proteomic_data.csv")
futimes <- read_csv("futimes.csv")
baseline_data <- read_csv("baseline_chars.csv")
```

Create joined dataset
```{r}
data_wide <- futimes %>%
  inner_join(cancer_data, by = "subjectid") %>%
  inner_join(proteomic_data, by = "subjectid") %>%
  inner_join(baseline_data, by = "subjectid") %>%
  mutate(
    cancer_futime = 2015 - v2date21_year, # followup time from visit 2 to 2015 (date of cancer status assessment)
    # create new variables for mm, cll, bnhl (types of cancer)
    mm = case_when(
      mye_lym1 == "Lymphoid (MM)" ~ 1,
      TRUE ~ 0), # in all other cases set equal to zero (justification: we have already filtered for !na lymphoid_inc, so cancer status is known)
    cll = case_when(
      mye_lym1 == "Lymphoid (CLL)" ~ 1,
      TRUE ~ 0), # in all other cases set equal to zero (justification: we have already filtered for !na lymphoid_inc, so cancer status is known)
    bnhl = case_when(
      mye_lym1 == "Lymphoid (B-NHL)" ~ 1,
      TRUE ~ 0) # in all other cases set equal to zero (justification: we have already filtered for !na lymphoid_inc, so cancer status is known)
  )

# first, calculate mean time intervals from visit 2 to visit 3
mean_v2_v3_gap <- mean(data_wide$v3_from_v2, na.rm = TRUE)
mean_v2_v5_gap <- mean(data_wide$v5_from_v2, na.rm = TRUE)

data_wide <- data_wide %>%
  rowwise() %>%
  mutate(
    event_from_v2 = case_when(
      !is.na(lyminc_t_v2) ~ lyminc_t_v2,
      !is.na(lyminc_t_v3) ~ lyminc_t_v3 + mean_v2_v3_gap,
      !is.na(lyminc_t_v5) ~ lyminc_t_v5 + mean_v2_v5_gap
    ),
    event_from_v3 = event_from_v2 - v3_from_v2,
    event_from_v5 = event_from_v2 - v5_from_v2
  ) %>%
  filter(event_from_v2 > 0) # only keep those with event time > 0 (ie diagnosed after visit 2)

```



Pivot longer
```{r}
data_long <- data_wide %>%
  pivot_longer(
    cols = matches("^v[235]_SeqId_\\d+_\\d+$"),
    names_to = c("visit", "protein"),
    names_pattern = "^v(\\d)_(SeqId_\\d+_\\d+)$",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = protein,
    values_from = value
  )
```


Add event_from_visit variable
```{r}
data_long_visittimes <- data_long %>%
  mutate(visit_time = case_when(
    visit == "2" ~ 0,
    visit == "3" ~ v3_from_v2,
    visit == "5" ~ v5_from_v2
  )) %>%
  group_by(subjectid) %>%
  arrange(visit_time) %>%
  mutate(event_from_visit = event_from_v2 - visit_time) %>% # calculate time from current visit to event/censoring
  ungroup()
```


Create event flag (set to 1 at the row where event occurs, 0 otherwise)

Delete proteomic measurements after event time
```{r}
data_long_filtered <- data_long_visittimes %>%
  filter(event_from_visit > 0) # remove rows (proteomic data) after cancer diagnosis 
```



## Create protein velocity variable 

```{r}
all_proteins_outcomes <- data_long_filtered %>%
  select(subjectid, bmi01, v1age01, center, gender, lymphoid_inc, mm, cll, bnhl, visit, visit_time, event_from_v2, SeqId_2665_26, SeqId_16322_10, SeqId_3292_75, SeqId_7849_3, SeqId_2704_74, SeqId_3291_30, SeqId_6574_11, SeqId_16915_153, SeqId_3291_30, SeqId_3487_32, SeqId_17346_61, SeqId_7266_4) %>% 
  filter(if_any(starts_with("SeqId"), ~!is.na(.)))  


```

Create lagged variables, calculate protein velocity (PV) and protein value at most recent visit (PR)

Function to prepare **standardised** tdc data
```{r}
prepare_std_tdc_data <- function(df, protein_col, event_col) {
  df %>%
    arrange(subjectid, visit) %>%
    group_by(subjectid) %>%
    mutate(
      # Create lagged variables for protein value and time
      protein_prev = lag(.data[[protein_col]], n = 1),
      time_prev = lag(visit_time, n = 1),
      
      # Protein velocity (PV)
      PV_calc = (.data[[protein_col]] - protein_prev) / (visit_time - time_prev),
      PV = case_when(is.na(PV_calc) ~ 0, TRUE ~ PV_calc),
      
      # Protein values
      PR = .data[[protein_col]],
      PB = first(.data[[protein_col]]),
      PL = last(.data[[protein_col]]),
      
      # Binary event indicator (per row)
      event_ind = ifelse(.data[[event_col]] == 1, 1, 0)
    ) %>%
    ungroup %>%
    
    # standardise the variables
    mutate(
      # Calculate means and SDs of the entire columns and standardize
      PB_std = (PB - mean(PB, na.rm = TRUE)) / sd(PB, na.rm = TRUE),
      PR_std = (PR - mean(PR, na.rm = TRUE)) / sd(PR, na.rm = TRUE),
      PL_std = (PL - mean(PL, na.rm = TRUE)) / sd(PL, na.rm = TRUE),
      # standardise PV based on non-zero values 
      mean_PV_non_zero = mean(PV[PV != 0], na.rm = TRUE),
      sd_PV_non_zero = sd(PV[PV != 0], na.rm = TRUE),
      
      # Use a conditional statement to standardize PV - keep it at zero when it was exactly zero
      PV_std = case_when(
        PV == 0 ~ 0, # If PV is zero, keep it zero
        TRUE ~ (PV - mean_PV_non_zero) / sd_PV_non_zero # Otherwise, standardize using non-zero values
      )
    ) %>%
    
    # survival formatting
    group_by(subjectid) %>%
    arrange(subjectid, visit_time) %>%
    mutate(
      tstart = visit_time, # define start time of interval as current visit time (e.g. v2 = 0, v3 = 3, v5 = 20)
      tstop = lead(visit_time, default = unique(event_from_v2)),  # usually set the stop time to the next visit (lead(visit_time). but if it's the last visit (ie before censoring/event) then it's the event_time
      
      # event indicator -- event = 1 if tstop is the event time and the event occurred, 0 otherwise
      event = ifelse(tstop == unique(event_from_v2) & event_ind == 1, 1, 0)
      )  %>%
    ungroup()
  
}

```


```{r}
data_std_test <- prepare_std_tdc_data(all_proteins_outcomes, "SeqId_2665_26", "mm")
```

## Function to fit tdc models 

```{r}
# function to fit tdc models
fit_std_tdc_models <- function(protein_col, outcome_col) {
  # prepare data 
  std_tdc_data <- prepare_std_tdc_data(all_proteins_outcomes, protein_col, outcome_col)
  
  # fit cox model for baseline 
  cox_model_baseline <- coxph(Surv(tstart, tstop, event) ~ PB_std + bmi01 + v1age01 + gender + center, data = std_tdc_data)
  
  # fit cox model for baseline and velocity 
  cox_model_baselinevelocity <- coxph(Surv(tstart, tstop, event) ~ PB_std + PV_std + bmi01+ v1age01 + gender + center, data = std_tdc_data)
  
  # return tidy results for each model
  bind_rows(
    tidy(cox_model_baseline) %>% 
      mutate(model = "baseline"),
    tidy(cox_model_baselinevelocity) %>%
      mutate(model = "baseline-velocity")
  ) %>%
    mutate(protein = protein_col, outcome = outcome_col)
}
```

Test
```{r}
test_model_std <- fit_std_tdc_models("SeqId_2665_26", "mm")
```

## Automate over protein-outcome combinations of interest
Loop over all protein-outcome combinations
```{r}
df_protein_mapping <- tibble::tibble(
  protein_seqid = c("SeqId_2665_26", "SeqId_16322_10", "SeqId_3292_75", "SeqId_7849_3", "SeqId_2704_74", "SeqId_3291_30", "SeqId_6574_11", "SeqId_16915_153", "SeqId_3291_30", "SeqId_3487_32", "SeqId_17346_61", "SeqId_7266_4"),
  outcome = c("mm", "mm", "mm", "mm", "mm", "cll", "cll", "cll", "bnhl", "bnhl", "bnhl", "bnhl")
)

# run both models for each mapping 
results_std_singleprotein_alloutcomes <- df_protein_mapping %>%
  pmap_dfr(~fit_std_tdc_models(..1, ..2))  # ..1 = protein_col, ..2 = outcome
```

Create summary table

# create summary table 
single_proteins_summary_std <- results_std_singleprotein_alloutcomes %>%
  filter(term %in% c("PB_std", "PR_std", "PV_std", "PL_std")) %>%
  select(protein, outcome, model, term,estimate, p.value, std.error) %>%
  pivot_wider(names_from = c(model, term), values_from = c(estimate, p.value, std.error)) %>% 
  mutate(
    across(contains("estimate"), ~ round(.x, 3)),
    across(contains("std.error"), ~ round(.x, 3)),
    across(contains("p.value"), ~ ifelse(.x < 0.001, 
                                         formatC(.x, format = "e", digits = 2), 
                                         formatC(.x, format = "f", digits = 3)))
  )

single_proteins_summary_std


```{r}
allmodels_sp_std_summary <- results_std_singleprotein_alloutcomes %>%
  filter(term %in% c("PB_std", "PV_std")) %>%
  select(protein, outcome, model, term, estimate, p.value, std.error) %>%
  mutate(
    HR = exp(estimate), # exponentiated log-HR
    CI_lower = exp(estimate - 1.96 * std.error),
    CI_upper = exp(estimate + 1.96 * std.error),
    CI = paste0("[", round(CI_lower, 3), ", ", round(CI_upper, 3), "]"),
    estimate = round(estimate, 3),       # log-HR
    HR = round(HR, 3),                   # HR
    std.error = round(std.error, 3),     # SE of log-HR
    signif = case_when(                  # significance stars
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    ),
    p.value = ifelse(p.value < 0.001, 
                     formatC(p.value, format = "e", digits = 2), 
                     formatC(p.value, format = "f", digits = 3))
  ) %>%
  select(-CI_lower, -CI_upper) 
allmodels_sp_std_summary
```

Write to csv

write_csv(single_proteins_summary_std, "tdc_allmodels_sp_summary_std.csv")


```{r}
write_csv(allmodels_sp_std_summary, "tdc_allmodels_sp_summary_std.csv")
```
